---
title: "In-Class Exercise 2"
editor: visual
---

## Getting Started

```{r}
pacman::p_load(sf, tidyverse, tmap, spdep, funModeling)
```

## Importing Geospatial Data

```{r}
#| eval: false
wp <- st_read(dsn = "geodata",
              layer = "geo_export",
              crs = 4326) %>%
  filter(clean_coun == "Nigeria")
```

### Importing water point geospatial data

```{r}
#| eval: false
write_rds(wp, "geodata/wp_nga.rds")
```

### Importing Nigeria LGA boundary data

```{r}
#| eval: false
nga <- st_read(dsn = "geodata",
               layer = "NGA",
               crs = 4326)
```

## Data Wrangling

### Recoding NA values into string

```{r}
#| eval: false
wp_nga <- read_rds("geodata/wp_nga.rds") %>%
    mutate(status_cle = replace_na(status_cle, "Unknown"))
```

### EDA

```{r}
#| eval: false
freq(data=wp_nga, 
     input = 'status_cle')
```

## Extracting Water Point Data

### Extracting functional water point

```{r}
#| eval: false
wpt_functional <- wp_nga %>%
  filter(status_cle %in%
           c("Functional", 
             "Functional but not in use",
             "Functional but needs repair"))
```

```{r}
#| eval: false
freq(data=wpt_functional, 
     input = 'status_cle')
```

### Extracting non-functional water point

```{r}
#| eval: false
wpt_nonfunctional <- wp_nga %>%
  filter(status_cle %in%
           c("Abandoned/Decommissioned", 
             "Abandoned",
             "Non-Functional",
             "Non functional due to dry season",
             "Non-Functional due to dry season"))
```

```{r}
#| eval: false
freq(data=wpt_nonfunctional, 
     input = 'status_cle')
```

### Extracting water point with Unknown class

```{r}
#| eval: false
wpt_unknown <- wp_nga %>%
  filter(status_cle == "Unknown")
```

## Performing Point-in-Polygon Count

```{r}
#| eval: false
nga_wp <- nga %>% 
  mutate(`total wpt` = lengths(
    st_intersects(nga, wp_nga))) %>%
  mutate(`wpt functional` = lengths(
    st_intersects(nga, wpt_functional))) %>%
  mutate(`wpt non-functional` = lengths(
    st_intersects(nga, wpt_nonfunctional))) %>%
  mutate(`wpt unknown` = lengths(
    st_intersects(nga, wpt_unknown)))
```

## Saving the Analytical Data Table

```{r}
#| eval: false
nga_wp <- nga_wp %>%
  mutate(pct_functional = `wpt functional`/`total wpt`) %>%
  mutate(`pct_non-functional` = `wpt non-functional`/`total wpt`)
```

```{r}
#| eval: false
write_rds(nga_wp, "geodata/nga_wp.rds")
```

## Visualising the spatial distribution of water points

```{r}
#| fig-width: 14
#| fig-height: 12
nga_wp <- read_rds("geodata/nga_wp.rds")

total <- qtm(nga_wp, "total wpt")+
  tm_layout(legend.outside = FALSE,
            legend.stack = "vertical",
            legend.text.size =0.3,
            legend.title.size=0.4)

wp_functional <- qtm(nga_wp, "wpt functional")+
  tm_layout(legend.outside = FALSE,
            legend.stack = "vertical",
            legend.text.size =0.3,
            legend.title.size=0.4)

wp_nonfunctional <- qtm(nga_wp, "wpt non-functional")+
  tm_layout(legend.outside = FALSE,
            legend.stack = "vertical",
            legend.text.size =0.3,
            legend.title.size=0.4)

unknown <- qtm(nga_wp, "wpt unknown")+
  tm_layout(legend.outside = FALSE,
            legend.stack = "vertical",
            legend.text.size =0.3,
            legend.title.size=0.4)

tmap_arrange(total, wp_functional, wp_nonfunctional, unknown, asp=1, ncol=2)
```
